---
title: "hw4_exercise"
output: html_document
date: "2023-02-05"
---

## Problem 1 

The goal here is to confirm with numerical experiments the asymptotic normality of the LAD estimator.

```{r}
sample_size <- c(10, 20, 50, 100, 200, 500)
R=1000

record_int_slope <- function(R, n) {
  inter <- c()
  slope <- c()
  for (i in 1:R) {
    x <- rexp(n,1)
    eps <- rt(n, 3)
    y <- -4 + 3*x + eps
    fit <- rq(y ~ x)
    inter[i] <- fit$coefficients[1]
    slope[i] <- fit$coefficients[2]
  }
  return(list(inter, slope))
}

for (i in 1:length(sample_size)) {
  hist(record_int_slope(R, sample_size[i])[[1]])
  hist(record_int_slope(R, sample_size[i])[[2]])
}

```

From above, we see that as n increases, the curve of the historgram of both intercept and the slop are more towards a bell curve. 

## Problem 2 

Consider the 

```{r}
require(car)
dat <- Florida
# plot the votes for Buchanan versus the total number of votes.
plot(BUCHANAN ~ Total, data = dat)
dat1 <- dat[-which(rownames(dat)=="PALM.BEACH"), ]
fit <- lm(BUCHANAN ~ Total, data = dat)
new <- data.frame(Total = dat[which(rownames(dat)=="PALM.BEACH"), "Total"]) # one number
predict(fit, new, interval = 'prediction', level = 0.99) # predicts new response

# another way 
#new <- dat[which(rownames(dat)=="PALM.BEACH"),]
#predict(fit, new, interval = 'prediction')
```

From above using the data without PALM.BEACH, we get a least square fit, which predict that PALM.BEACH will have 1052.449 vote for BUCHANAN. In reality, it has 3407 for BUCHANAN.

```{r}
dat2 = log(dat)
plot(BUCHANAN ~ Total, data = dat2)
dat2.1 <- dat2[-which(rownames(dat2)=="PALM.BEACH"), ]
fit2 <- lm(BUCHANAN ~ Total, data = dat2.1)
abline(fit2, lwd = 3, col = 2)
new <- data.frame(Total = dat2[which(rownames(dat2)=="PALM.BEACH"), "Total"]) # one number
predict(fit2, new, interval = 'prediction', level = 0.99) # predicts new response

```

The fit is getting better. 


