---
title: "hw6_exercise"
output: html_document
date: "2023-02-28"
---

## Problem 1 

### Part A
Write a function `piecewiseConstant(x, y, q = 10, plot = TRUE)` taking in a one dimensional
predictor variable `x` with values in $[0, 1]$ and a response `y`, and fits a piecewise constant model
(by least squares) on q intervals of equal length partitioning the unit interval, with the option
of producing a scatterplot with the fit overlaid.

### Part B
Apply your function to some dataset of your choice, either synthetic or real. (If you choose a
real dataset, then provide the source.) Produce a single scatterplot, with lines corresponding
to the fit with `q = 5 (blue), q = 10 (green), and q = 20 (red)`. Add a legend.

```{r}

f = function(x) (1 + 10*x - 5*x^2)*sin(10*x)
n = 200
x = runif(n, 0, 1)
x = sort(x) # for convenience (see below)
y = f(x) + rnorm(n)
plot(x, y)
t = seq(0, 1, len = 1000)
lines(t, f(t), lwd = 2, lty = 2)

piecewiseConstant <- function(x, y, q = 10, plot = TRUE) {
  
  all.breaks = (0:q)/q
  inner.breaks = (1:(q-1))/q
  val = numeric(q)
  for (k in 1:q){
	  ind = (all.breaks[k] < x)&(x <= all.breaks[k+1])
	  val[k] = mean(y[ind])
  }
  fhat = stepfun(inner.breaks, val)
  
  if (plot == TRUE) {
    plot(x, y)
    abline(v = all.breaks, lty = 3)
    lines(fhat, col = 2, lwd = 2, do.points = FALSE)
  }
}

piecewiseConstant(x, y, q= 10, plot= TRUE)


```

```{r}
piecewiseConstant2 <- function(x, y, q, plot = TRUE) {
  
  model <- vector(mode='list', length=3)
  
  for(i in 1:length(q)) {
    # all.breaks = (0:q[i])/q[i]
    inner.breaks = (1:(q[i]-1))/q[i]
    val = numeric(q[i])
    for (k in 1:q[i]){
	    ind = (all.breaks[k] < x)&(x <= all.breaks[k+1])
	    val[k] = mean(y[ind])
    }
    fhat = stepfun(inner.breaks, val)
    model[[i]] <- fhat
  }
  
  if (plot == TRUE) {
    plot(x, y)
    # abline(v = all.breaks, lty = 3)
    for (i in 1:length(q)){
      lines(model[[i]], col = i, lwd = 2, do.points = FALSE)
    }
    legend("topleft", legend=c("d=5", "d=10", "d=20"), col=c(1,2,3), lwd=2)
  }
}

piecewiseConstant2(x, y, q=c(5, 10, 20), plot= TRUE)

```


## Problem 2 

### Part A
Fit a simple logistic model. Explain why the performance is bound to be mediocre.

### Part B
Fit a higher order polynomial logistic model. Consider several variants: degree 2 without interactions, degree 2 with interactions, and higher degree models. Track the average performance by averaging over R = 20 repeats. Produce a plot or two summarizing the results.

```{r}
library(MASS)
n = 50
x <- as.data.frame(mvrnorm(n, c(0,0), diag(2)))
gen <- function(s) {
  y <- rbinom(1, size = 1, exp(-norm(s, type = "2")))
  return(y)
}
y <- apply(x, MARGIN = 1, FUN = gen)



fit1 = glm(y ~ x[,1] + x[, 2], family = binomial(link = "logit"))
summary(fit1) # none of the estimator is significant 

# degree 2 without interactions
fit2 = glm(y ~ x[,1] + x[, 2] + I(x[,1]^2) + I(x[,1]^2), family = binomial(link = "logit"))
summary(fit2)

# degree 2 with interactions
fit3 = glm(y ~ x[,1] + x[, 2] + I(x[,1]^2) + I(x[,1]^2) + x[,1] * x[,2] + I(x[,1]^2) * I(x[,1]^2), family = binomial(link = "logit"))
summary(fit3)

# higher degree models
fit4 = glm(y ~ x[,1] + x[, 2] + I(x[,1]^10) + I(x[,1]^10), family = binomial(link = "logit"))
summary(fit3)

# Does not fit very well 
```



